---
title: "Summary of Massachusetts Tax Override Data"
author: Gang Chen
Email: gchen3@albany.edu
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: html
---

## Main data sources
Mass DLS databank
https://www.mass.gov/collections/DLS-databank-reports?topicid=91816

## Loading and cleaning data
1. Read xls data to R
2. Identify the variables and observations
3. Merge all years data
4. Match municipal code with municipal name

```{r data}
#| echo: false
#| include: false

library(tidyverse)
library(foreign)
library(readxl)
library(here)
library(gt)

filepath <- getwd()
setwd("C:/Users/Gang Chen/OneDrive - University at Albany - SUNY/r-projects/mass-override")




read_municipal_debt <- function(file_name) {
  
debt <- read_excel(
  path = here::here("data", file_name),
  sheet = "Debt",
  range = cell_limits(c(7, 2), c(NA, NA)),
  col_names = FALSE)

column_names <- debt[1,]

debt <- read_excel(
  path = here::here("data", file_name),
  sheet = "Debt",
  range = cell_limits(c(9, 2), c(NA, NA)),
  col_names = FALSE)

colnames(debt) <- as.character(column_names)

if ("DOR Code" %in% colnames(debt)) {
    debt$`DOR Code` <- as.numeric(debt$`DOR Code`)
  }

return(debt)
}

debt_data_2022 <- read_municipal_debt("municipaldebt2022.xlsx")
debt_data_1019 <- read_municipal_debt("municipaldebt1019.xls")
debt_data_0009 <- read_municipal_debt("municipaldebt0009.xls")
debt_data_9199 <- read_municipal_debt("municipaldebt9199.xls")

debt_all_years <- bind_rows(debt_data_9199, debt_data_0009, debt_data_1019, debt_data_2022)

debt_all_clean <- debt_all_years |>
  filter(!is.na(`DOR Code`)) |>
  rename_with(~ str_replace_all(.x, " ", "_") |>
                str_replace_all("%", "percent") |>  
                tolower())  

muni_name_code <- read_excel(
  path = here::here("data", "municipaldebt2022.xlsx"),
  sheet = "Debt",
  range = cell_limits(c(7, 1), c(NA, 2)),
  col_names = TRUE) 

muni_name_code$`DOR Code` <- as.numeric(muni_name_code$`DOR Code`)

muni_name_code <- muni_name_code |>
  filter(!is.na(Municipality)) |>
  rename_with(~ str_replace_all(.x, " ", "_") |>
                tolower())

debt_all_clean <- debt_all_clean |>
  left_join(muni_name_code, by = c("dor_code" = "dor_code"))

saveRDS(debt_all_clean, here::here("data", "debt_all_clean.RData"))

```

## Summarize debt data and show tables
1. Write a function that shows the average number per year 
2. Write a function that shows the number for Acton per year

```{r}
#| echo: false
#| include: false

debt_all_clean <- readRDS(here::here("data", "debt_all_clean.RData"))

names(debt_all_clean)

get_debt_table <- function(var = debt_service_as_percent_of_budget) {

yearly_data <- debt_all_clean |>
    group_by(fy) |>
    summarize(average = mean({{var}}, na.rm = TRUE),
              count = n(), .groups = 'drop') 

return(yearly_data)

}
  
get_debt_table(debt_service_as_percent_of_budget)

get_acton_table <- function(var = debt_service_as_percent_of_budget) {

acton_data <- debt_all_clean |>
    filter(municipality == "Acton") |>
    select(municipality, fy, {{var}})
  
return(acton_data)
}

get_acton_table(debt_service_as_percent_of_budget)

```
## Produce Tables and figures
```{r}
#| echo: false
#| include: false

debt_service_year <- get_debt_table(debt_service_as_percent_of_budget)
debt_service_acton <- get_acton_table(debt_service_as_percent_of_budget)
debt_service_compare <- debt_service_year |> 
  left_join(debt_service_acton, by = "fy") |>
  select(-count, -municipality) |>
  rename(Acton = debt_service_as_percent_of_budget)

debt_outstanding_year <- get_debt_table(debt_as_percent_of_eqv)
debt_outstanding_acton <- get_acton_table(debt_as_percent_of_eqv)
debt_outstanding_compare <- debt_outstanding_year |> 
  left_join(debt_outstanding_acton, by = "fy") |>
  select(-municipality) |>
  rename(Acton = debt_as_percent_of_eqv)
  
debt_outstanding_compare


```
### Table 1 Debt Service as a Percent of Budget, Acton vs MA average, by year
```{r}
#| echo: false

debt_service_compare |> 
  gt() |>
  tab_header(
    title = md("**Debt Service as Percent of Budget, Acton vs MA Average, 1991-2022**")
  ) |>
  tab_source_note(
    source_note = "Source: Department of Local Services Databank."
  ) |>
  cols_label(
    fy = md("Year"),
    Acton = md("Acton (% of budget)"),
    average = md("MA Average (% of budget)")) |>
  fmt_number(
    columns = c(Acton, average),
    decimals = 2
  )
```
### Table 2 Debt Outstanding divided by Equalized Valuations, Acton vs MA average, by year
```{r}
#| echo: false

debt_outstanding_compare |> 
  select(-count) |>
  gt() |>
  tab_header(
    title = md("**Debt Outstanding divided by Equalized Valuations, Acton vs MA Average, 1991-2022**")
  ) |>
  tab_source_note(
    source_note = "Source: Department of Local Services Databank."
  ) |>
  cols_label(
    fy = md("Year"),
    Acton = md("Acton (debt/evaluation)"),
    average = md("MA Average (debt/evaluation)")) |>
  fmt_number(
    columns = c(Acton, average),
    decimals = 2
  )
```
### Figure 1 Debt Service as a Percent of Budget, Acton vs MA average, by year
```{r}
#| echo: false

debt_service_long <- debt_service_compare |>
  rename(MA_average = average) |>
  pivot_longer(cols = c(Acton, MA_average), 
               names_to = "Location", 
               values_to = "Debt_Service_Percent")

ggplot(debt_service_long, aes(x = fy, y = Debt_Service_Percent, color = Location, linetype = Location)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Debt Service as Percent of Budget, Acton vs MA Average, 1991-2022",
       x = "Year",
       y = "Debt Service as Percent of Budget (%)") +
  theme_minimal() +
  scale_color_manual(values = c("Acton" = "black", "MA_average" = "black")) +
  scale_linetype_manual(values = c("Acton" = "solid", "MA_average" = "dashed")) +
  theme(legend.position = "bottom",         
        legend.title = element_blank())

```
### Figure 2 Debt Outstanding divided by Equalized Valuations, Acton vs MA average, by year
```{r}
#| echo: false

debt_outstanding_long <- debt_outstanding_compare |>
  rename(MA_average = average) |>
  pivot_longer(cols = c(Acton, MA_average), 
               names_to = "Location", 
               values_to = "Debt_outstanding")

ggplot(debt_outstanding_long, aes(x = fy, y = Debt_outstanding, color = Location, linetype = Location)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Debt Outstanding divided by Equalized Valuations, Acton vs MA average, 1991-2022",
       x = "Year",
       y = "debt/evaluation") +
  theme_minimal() +
  scale_color_manual(values = c("Acton" = "black", "MA_average" = "black")) +
  scale_linetype_manual(values = c("Acton" = "solid", "MA_average" = "dashed")) +
  theme(legend.position = "bottom",         
        legend.title = element_blank())

```
